---
title: "Getting Started with lexiExpand"
author: "Gopi Nath Vajpai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lexiExpand}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Why lexiExpand?

Dictionary-based text analysis is one of the most widely used methods in
computational social science, finance, and political science. Researchers
count how often specific words appear in documents to measure latent
concepts — *uncertainty*, *innovation*, *affect* — without training a
complex model from scratch.

The catch: **dictionary coverage**. Miss the word "automobile" in your
*Transportation* dictionary and every document that uses it is silently
under-counted. Closing these gaps traditionally required either painful
manual synonym searches or training a full Word2Vec / GloVe model.

`lexiExpand` automates that step. You give it a small **seed dictionary**;
it uses pre-trained GloVe embeddings to find the nearest words in semantic
space and walks you through an interactive accept/reject review. The result
is a richer, more reproducible dictionary in minutes.

---

## Installation

```{r install}
# From a local source tarball
install.packages("lexiExpand_0.2.0.tar.gz", repos = NULL, type = "source")

# Required dependencies (installed automatically)
# cli, data.table, rappdirs

# Optional — needed only for the quanteda export format
install.packages("quanteda")

# Optional — alternative vector source
install.packages("textdata")
```

---

## Step 1 — Download word vectors (once)

`lexiExpand` ships without bundled vectors to keep the package small and
CRAN-compatible. On first use, call `download_vectors()` to fetch the
Stanford GloVe 6B model and cache it on disk. This step happens **once
per machine**.

```{r download}
library(lexiExpand)

# Downloads ~822 MB zip, extracts the 50-dimension file (~160 MB),
# then deletes the zip. Subsequent calls detect the cache and skip.
download_vectors("glove.6B.50d")
#> ! Downloading the GloVe 6B source archive (~822 MB).
#>   This is a one-time download; the zip is deleted after extraction.
#> i URL: https://downloads.cs.stanford.edu/nlp/data/glove.6B.zip
#> Extracting glove.6B.50d.txt from archive...
#> Deleted source zip to save disk space.
#> v Done! Vectors cached at:
#>   ~/.local/share/lexiExpand/glove.6B.50d.txt
```

The cache lives in your OS user-data directory (managed by `rappdirs`):

| Platform | Default path |
|----------|-------------|
| Linux    | `~/.local/share/lexiExpand/` |
| macOS    | `~/Library/Application Support/lexiExpand/` |
| Windows  | `C:\Users\<you>\AppData\Local\lexiExpand\` |

You can override this with the `cache_dir` argument on any function.
If the download times out, increase the limit:

```{r timeout}
download_vectors("glove.6B.50d", timeout = 7200)  # 2 hours
```

**Alternative:** use the `textdata` package as the vector source — it
handles its own consent-and-cache flow:

```{r textdata-source}
# install.packages("textdata")
vecs <- load_vectors_textdata(dimensions = 300, vocab_size = 50000)
```

---

## Step 2 — Expand a seed dictionary

### 2a. Interactive wizard (recommended for exploratory work)

```{r expand-interactive}
result <- expand_dict(
  seed      = c("sad", "angry"),
  n         = 20,          # candidates to surface per seed word
  threshold = 0.70         # minimum cosine similarity to show
)
```

The wizard displays each candidate with its similarity score. Use the
keyboard to review:

```
── Semantic Dictionary Expander — Review Wizard ──────────────────────────────
Seed words: "sad" and "angry"
20 candidates to review, sorted by similarity.
Keys: y accept · n skip · a accept all remaining · q quit

── [1/20] ─────────────────── similarity: 84%  |  seed: sad ──────────────────
miserable
  > y
✔ Added "miserable"
```

### 2b. Non-interactive / scripted mode

Set `interactive = FALSE` to get the raw ranked data frame — useful inside
RMarkdown documents, batch scripts, or automated pipelines.

```{r expand-noninteractive}
candidates <- expand_dict(
  seed        = c("sad", "angry"),
  n           = 20,
  threshold   = 0.65,
  interactive = FALSE
)

head(candidates)
#>          word similarity   seed pct_match
#> 1   miserable      0.843    sad       84%
#> 2     furious      0.821  angry       82%
#> 3     unhappy      0.791    sad       79%
#> 4   resentful      0.778  angry       78%
#> 5      gloomy      0.764    sad       76%
#> 6      bitter      0.751  angry       75%
```

### 2c. Centroid mode

Instead of finding neighbours for each seed separately, centroid mode
averages all seeds into a single vector and finds words nearest that point.
Better when your seeds form a coherent concept.

```{r centroid}
candidates_c <- expand_dict(
  seed        = c("sad", "angry"),
  seed_mode   = "centroid",    # default is "individual"
  n           = 20,
  threshold   = 0.65,
  interactive = FALSE
)
# $seed column is "centroid" for every row
```

### 2d. Re-using a loaded matrix (efficient repeated calls)

```{r reuse-vectors}
vecs <- load_vectors("glove.6B.50d")

neg_emotion  <- expand_dict(c("sad", "angry"),
                             vectors = vecs, interactive = FALSE)
pos_emotion  <- expand_dict(c("happy", "joyful"),
                             vectors = vecs, interactive = FALSE)
uncertainty  <- expand_dict(c("uncertain", "ambiguous", "unclear"),
                             vectors = vecs, interactive = FALSE)
```

---

## Step 3 — Expand a full lexicon data frame

`expand_lexicon()` accepts a two-column data frame (word + dimension label)
and expands every dimension in one call. This is the main entry point when
working with an established lexicon like the MDE dictionary from
[mdeinR](https://github.com/gvajpai/mdeinR).

```{r expand-lexicon}
my_lexicon <- data.frame(
  word      = c("aroma",     "flavor",    "texture",
                "delighted", "excited",   "moved",
                "welcoming", "attentive", "friendly"),
  dimension = c("sensory",   "sensory",   "sensory",
                "affect",    "affect",    "affect",
                "social",    "social",    "social")
)

candidates <- expand_lexicon(
  lexicon   = my_lexicon,
  n         = 15,
  threshold = 0.68,
  vectors   = vecs
)

# Results have a $dimension column
table(candidates$dimension)
#> affect sensory  social
#>     12      14      11
```

If your column names differ from `word`/`dimension`:

```{r custom-cols}
expand_lexicon(my_lexicon,
               word_col = "term",
               dim_col  = "group",
               vectors  = vecs)
```

---

## Step 4 — Export to your analysis tool

### Plain R list

```{r export-list}
dict_list <- export_dict(
  result = candidates,          # data.frame from expand_dict / expand_lexicon
  seed   = c("sad", "angry"),
  name   = "negative_emotion",
  format = "list"
)

dict_list
#> $negative_emotion
#>  [1] "sad"       "angry"     "miserable" "furious"   ...

saveRDS(dict_list, "dictionaries/negative_emotion.rds")
```

### quanteda dictionary

```{r export-quanteda}
library(quanteda)

neg_dict <- export_dict(
  result = result,              # lexiexpand_result from interactive wizard
  name   = "negative_emotion",
  format = "quanteda"
)

# Score a corpus
corp   <- corpus(c(
  doc1 = "She felt sad and miserable after the news.",
  doc2 = "He was furious but tried to stay calm.",
  doc3 = "The report showed stable and optimistic results."
))
scored <- dfm(tokens(corp)) |>
  dfm_lookup(dictionary = neg_dict) |>
  convert(to = "data.frame")

scored
#>   doc_id  negative_emotion
#> 1   doc1                 2
#> 2   doc2                 1
#> 3   doc3                 0
```

### Two-column data frame (tidytext / CSV)

```{r export-df}
dict_df <- export_dict(
  result = candidates,
  seed   = c("sad", "angry"),
  name   = "negative_emotion",
  format = "data.frame"
)

write.csv(dict_df, "dictionaries/negative_emotion.csv", row.names = FALSE)
```

---

## Full scripted pipeline (MDE example)

```{r full-pipeline}
library(lexiExpand)
library(mdeinR)     # provides mde_words (word + dimension columns)

# 1. Setup (first run only)
download_vectors("glove.6B.50d")

# 2. Load vectors once
vecs <- load_vectors("glove.6B.50d", vocab_size = 100000L)

# 3. Expand all MDE dimensions from the lexicon data frame
candidates_mde <- expand_lexicon(
  lexicon   = mde_words,
  n         = 25L,
  threshold = 0.68,
  seed_mode = "individual",
  vectors   = vecs
)

# 4. Keep top-10 per dimension
library(data.table)
top10 <- as.data.table(candidates_mde)[
  , head(.SD, 10L), by = dimension
]

# 5. Build augmented dictionary list
mde_aug <- lapply(split(mde_words$word, mde_words$dimension), identity)
for (dim in unique(top10$dimension)) {
  mde_aug[[dim]] <- unique(c(mde_aug[[dim]], top10[dimension == dim, word]))
}

# 6. Export to quanteda and save
library(quanteda)
full_dict <- quanteda::dictionary(mde_aug)
saveRDS(full_dict, "mde_dictionary_augmented.rds")

# 7. Score reviews
reviews <- corpus(c(
  r1 = "The aroma was captivating and the presentation absolutely stunning.",
  r2 = "Staff were genuinely warm — we felt truly part of a community.",
  r3 = "We learned about fermentation — a fascinating culinary journey.",
  r4 = "Very average. Nothing remarkable about the food or atmosphere."
))

dfm(tokens(reviews)) |>
  dfm_lookup(dictionary = full_dict) |>
  convert(to = "data.frame")
#>   doc_id sensory affect social intellectual behavioral
#> 1     r1       3      1      0            0          0
#> 2     r2       0      1      3            0          0
#> 3     r3       0      1      0            3          0
#> 4     r4       0      0      0            0          0
```

---

## Tips and best practices

**Choose your threshold carefully.** A threshold of 0.65–0.70 gives a broad
semantic neighbourhood; 0.80+ restricts to near-synonyms. For publication
work, review candidates manually (`interactive = TRUE`) rather than
accepting all.

**Watch for antonyms.** Word embedding spaces cluster by *distributional
context*, not polarity. "Sad" and "glad" appear in similar contexts, so
"glad" may appear as a high-similarity candidate for a negative emotion
seed. Always review candidates before accepting.

**Higher dimensions = better neighbours (usually).** The 50-dimension model
loads fast and works well for common English words. For domain-specific
vocabulary (hospitality, legal, financial), try `"glove.6B.100d"` or
`"glove.6B.200d"`.

**Reproducibility.** Save both the final dictionary *and* the full candidate
table so you can document which words were considered:

```{r reproducibility}
write.csv(candidates_mde,
          "appendix_all_candidates.csv",
          row.names = FALSE)
```
